Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-12-16 summary.Rmd
fixed a few citations
just syncing some stuff up

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabilistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  community detection and parameter estimation. In particular, we note that the
  PABM can be represented as latent positions such that points within the same 
  community lie on a subspace, and the subspaces that represent each 
  community are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our community 
  detection and parameter estimation methods for the PABM. 
---

@@

Statistical analysis on graphs or networks often involves the partitioning of 
a graph into disconnected subgraphs or clusters. This is often motivated by the 
belief that there exist underlying and unobserved communities to which each 
vertex of the graph belongs, and edges between pairs of vertices are determined 
by drawing from a probability distribution based on the community relationships 
between each pair. The goal of this analysis then is population community 
detection, or the recovery of the true underlying community labels for each 
vertex, up to permutation (with some additional parameter estimation being of 
possible interest), assuming some probabilistic graph model. One such model is 
the Stochastic Block Model (SBM), first proposed by 
\citet{doi:10.1080/0022250X.1971.9989788}, which assumes that the edge 
probability from one vertex to another follows a Beronulli distribution with 
fixed probabilities for each pair of community labels. Other probabilistic 
graph models have been proposed and studied, such as the Degree-Corrected Block 
Model (DCBM), introduced by \citet{Karrer_2011} and the Mixed Membership 
Stochastic Block Model (MMSBM), introduced by  \citet*{NIPS2008_3578}. The 
Popularity Adjusted Block Model (PABM) was then introduced by 
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the SBM to 
address the heterogeneity of edge probabilities within and between communities 
while still maintaining distinct community structure.

@@

The Random Dot Product Graph (RDPG) model proposed by 
\citet*{10.1007/978-3-540-77004-6_11} is another graph model with Bernoulli edge 
probabilities. Under this model, each vertex of the graph can be represented by 
a point in a latent Euclidean space such that the edge probability between any 
pair of vertices is given by their corresponding dot product in the latent 
space, i.e., given a latent positions $x_1, ..., x_n \in \mathbb{R}^d$, the 
edge probability matrix is $P = X X^\top$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. The SBM is 
equivalent to a special case of the RDPG model in which all vertices of a 
given community share the same position in the latent space 
\cite{lyzinski2015perfect}. It has also been shown that similar 
probabilistic graph models, including both the MMSBM and DCBM, can be 
represented in this way \cite{rubindelanchy2017consistency} 
\cite{lyzinski2015perfect}. An analogous 
property exists for the PABM but not for the RDPG model but under the 
*Generalized* Random Dot Product Graph (GRDPG) model. This relationship will be 
explored in this paper and exploited to construct algorithms for community 
detection and parameter estimation for the PABM.

@@

\citet*{307cbeb9b1be48299388437423d94bf1}, who first proposed the PABM, used 
modularity maximization for clustering and parameter estimation. They were able 
to show that as the sample size increases, the proportion of misclassified 
community labels (up to permutation) goes to 0.

@@

Sparse subspace clustering is described as follows. First, an embedding of the 
affinity matrix $A$ is required, and for this, we use spectral decomposition 
using the first $K (K + 1) / 2$ and last $K (K - 1) / 2$ eigenvectors. Then a 
sparse linear regression model is fit to each row of the embedding data matrix 
using the remaining rows as predictors, the rows of which are stored in 
matrix $C$. Finally, an affinity matrix $B = |C| + |C^\top|$ is constructed, 
where $|\cdot|$ is applied entry-wise. If the data in the embedding lie 
perfectly in $K$ subspaces, then $B_{ij} = 0$ for $i, j$ not in the same 
subspace. If the embedding of $P$ is used, then this property holds, but it 
is true in general for the embedding of $A$, in which case a final clustering 
step is required. For these simulations, spectral clustering with the 
normalized Laplacian and Gaussian mixture models is used for this final step.

>>>>

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two random graph models, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  community detection and parameter estimation. In particular, we note that the
  PABM can be represented as latent positions such that points within the same 
  community lie on a subspace, and the subspaces that represent each 
  community are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our community 
  detection and parameter estimation methods for the PABM. 
---

@@

Statistical analysis on graphs or networks often involves the partitioning of 
a graph into disconnected subgraphs or clusters. This is often motivated by the 
belief that there exist underlying and unobserved communities to which each 
vertex of the graph belongs, and edges between pairs of vertices are determined 
by drawing from a probability distribution based on the community relationships 
between each pair. The goal of this analysis then is population community 
detection, or the recovery of the true underlying community labels for each 
vertex, up to permutation (with some additional parameter estimation being of 
possible interest), assuming some random graph model. One such model is 
the Stochastic Block Model (SBM), first proposed by 
\citet{doi:10.1080/0022250X.1971.9989788}, which assumes that the edge 
probability from one vertex to another follows a Beronulli distribution with 
fixed probabilities for each pair of community labels. Other random 
graph models have been proposed and studied, such as the Degree-Corrected Block 
Model (DCBM), introduced by \citet{Karrer_2011} and the Mixed Membership 
Stochastic Block Model (MMSBM), introduced by  \citet*{NIPS2008_3578}. The 
Popularity Adjusted Block Model (PABM) was then introduced by 
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the SBM to 
address the heterogeneity of edge probabilities within and between communities 
while still maintaining distinct community structure.

@@

The Random Dot Product Graph (RDPG) model proposed by 
\citet*{10.1007/978-3-540-77004-6_11} is another graph model with Bernoulli edge 
probabilities. Under this model, each vertex of the graph can be represented by 
a point in a latent Euclidean space such that the edge probability between any 
pair of vertices is given by their corresponding dot product in the latent 
space, i.e., given a latent positions $x_1, ..., x_n \in \mathbb{R}^d$, the 
edge probability matrix is $P = X X^\top$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. The SBM is 
equivalent to a special case of the RDPG model in which all vertices of a 
given community share the same position in the latent space 
\cite{lyzinski2014}. It has also been shown that similar 
random graph models, including both the MMSBM and DCBM, can be 
represented in this way \cite{rubindelanchy2017consistency} 
\cite{lyzinski2014}. An analogous 
property exists for the PABM but not for the RDPG model but under the 
*Generalized* Random Dot Product Graph (GRDPG) model. This relationship will be 
explored in this paper and exploited to construct algorithms for community 
detection and parameter estimation for the PABM.

@@

\citet*{307cbeb9b1be48299388437423d94bf1}, who first proposed the PABM, used 
modularity maximization for clustering and parameter estimation. They were able 
to show that as the sample size increases, the proportion of misclassified 
community labels (up to permutation) goes to 0. However, because modularity 
maximization is NP-hard, the authors of this paper used an approximation 
based on the Extreme Points algorithm \cite{le2016}.

@@

Sparse subspace clustering is described as follows. First, an embedding of the 
affinity matrix $A$ is required, and for this, we use spectral decomposition 
using the first $K (K + 1) / 2$ and last $K (K - 1) / 2$ eigenvectors. Then a 
sparse linear regression model is fit to each row of the embedding data matrix 
using the remaining rows as predictors, the rows of which are stored in 
matrix $C$. Finally, an affinity matrix $B = |C| + |C|^\top$ is constructed, 
where $|\cdot|$ is applied entry-wise. If the data in the embedding lie 
perfectly in $K$ subspaces, then $B_{ij} = 0$ for $i, j$ not in the same 
subspace. If the embedding of $P$ is used, then this property holds, but it 
is true in general for the embedding of $A$, in which case a final clustering 
step is required. For these simulations, spectral clustering with the 
normalized Laplacian and Gaussian mixture models is used for this final step.
